\documentclass[reqno]{article}
\usepackage{../format-doc}

\begin{document}
	\title{Maier-Saupe free energy in weak form}
	\author{Lucas Myers}
	\maketitle
	
	\section{Introduction}
	Here we will find a PDE describing the time evolution equation of the $Q$-tensor from thermodynamic effects, according to the Maier-Saupe free energy.
	Then we will discretize time according to a general finite difference scheme.
	After this, we will put the resulting space-dependent equations into weak form.
	The result will be non-linear, so we will have to use Newton's method to compute the solution to the finite difference scheme.
	
	\section{Maier-Saupe free energy and equations of motion}
	\subsection{Writing the free energy in terms of $Q_{ij}$}
	We begin by defining the tensor order parameter of the nematic system in terms of the probability distribution of the molecular orientation:
	\begin{equation} \label{eq:Q-def}
		Q_{ij} (\mathbf{x}) 
		= \int_{S^2} \left( \xi_i \xi_j - \tfrac13 \delta_{ij} \right)
		p(\mathbf{\xi} ; \mathbf{x}) d \mathbf{\xi}
	\end{equation}
	where $p(\mathbf{\xi} ; \mathbf{x})$ is the probability distribution of molecular orientation in local equilibrium at some temperature $T$ and position $\mathbf{x}$. 
	Note that this quantity is traceless and symmetric.
	Then the mean field free energy is given by:
	\begin{equation}
		F \left[ Q_{ij} \right] = H \left[ Q_{ij} \right] - T \Delta S
	\end{equation}
	where $H$ is the energy of the configuration, and $\Delta S$ is the entropy relative to the uniform distribution.
	We choose $H$ to be:
	\begin{equation}
		H\left[ Q_{ij} \right]
		= \int_{\Omega} \left\{ -\alpha Q_{ij} Q_{ji} + f_e 	\left( Q_{ij}, \partial_k Q_{ij} \right) \right\} d \mathbf{x}
	\end{equation}
	with $\alpha$ some interaction parameter and $f_e$ the elastic free energy density.
	The entropy is given by:
	\begin{equation} \label{eq:entropy-def}
		\Delta S 
		= -n k_B \int_{\Omega} \left(
		\int_{S^2} p(\mathbf{\xi} ; \mathbf{x}) 
		\log \left[ 4 \pi p( \mathbf{\xi} ; \mathbf{x} ) \right] d \mathbf{\xi} \right) d \mathbf{x}
	\end{equation}
	where n is the number density of molecules.
	Now, in general for a given $Q_{ij}$ there is no unique $p(\mathbf{\xi} ; \mathbf{x})$ given by \eqref{eq:Q-def}. 
	Hence, there is no unique $\Delta S$.
	To find the appropriate $\Delta S$ corresponding to some fixed $Q_{ij}$, we seek to maximize the entropy density for a fixed $Q_{ij}$ via the method of Lagrange multipliers. 
	This goes as follows:
	\begin{equation}
	\begin{split}
		\mathcal{L} [p]
		&= \Delta s [p] 
		+ \Lambda_{ij} \left[\int_{S^2} 
			\left(\xi_i \xi_j - \tfrac13 \delta_{ij}\right) p\left(\xi; \mathbf{x}\right) d\xi
			- Q_{ij} \right] \\
		&= \int_{S^2} p( \mathbf{\xi}) 
		\biggl(
		-n k_B\log \left[ 4\pi p(\mathbf{\xi}) \right]
		+ \Lambda_{ij} \left( \xi_i \xi_j - \tfrac13 \delta_{ij} \right)
		\biggr) d\mathbf{\xi} - \Lambda_{ij} Q_{ij}
	\end{split}
	\end{equation}
	where here $Q_{ij}$ is taken to be a constant which defines the constraint.
	Here we've taken the spatial dependence to be implicit, since each of these are local quantities, and we're minimizing them \textit{locally}.
	So, define a variation in $p$ given by:
	\begin{equation}
		p'(\mathbf{\xi}) 
		= p(\mathbf{\xi}) + \varepsilon \eta(\mathbf{\xi})
	\end{equation}
	Then we have that:
	\begin{equation}
	\begin{split}
		\frac{\delta \mathcal{L}}{\delta p}
		&= \left.\frac{d \mathcal{L}[p']}{d \varepsilon} \right|_{\varepsilon = 0} \\
		&= \left.\frac{d \mathcal{L} [p']}{d p'} \frac{d p'}{d \varepsilon} \right|_{\varepsilon = 0} \\
		&= \int_{S^2} \biggl(
		-n k_B\log \left[ 4\pi p(\mathbf{\xi}) \right]
		+ \Lambda_{ij} \left( \xi_i \xi_j - \tfrac13 \delta_{ij} \right)
		- n k_B
		\biggr) \eta(\mathbf{\xi}) d \mathbf{\xi}
	\end{split}
	\end{equation}
	Since this is for an arbitrary variation $\eta$, we must have that
	\begin{equation}
	-n k_B\log \left[ 4\pi p(\mathbf{\xi}) \right]
	+ \Lambda_{ij} \left( \xi_i \xi_j - \tfrac13 \delta_{ij} \right)
	- n k_B
	= 0
	\end{equation}
	Solving for $p(\mathbf{\xi})$ yields:
	\begin{equation}
		p(\mathbf{\xi}) 
		= \frac{1}{4 \pi}
		\exp \left[
		- \tfrac{1}{n k_B} \tfrac13 \Lambda_{ij} \delta_{ij} -  1
		\right]
		\exp\left[
		\tfrac{1}{n k_B} \Lambda_{ij} \xi_i \xi_j
		\right]
	\end{equation}
	However, $p(\mathbf{\xi})$ is a probability distribution, so we need to normalize it over the domain.
	When we do this, the constant factors out front cancel.
	Further, since $\Lambda_{ij}$ is just an arbitrary factor that we use to calculate $p(\xi)$, we may rescale it by $n k_B$.
	Doing all this yields:
	\begin{equation} \label{eq:p-expr}
		p( \mathbf{\xi} )
		= \frac{\exp\left[ \Lambda_{ij} \xi_i \xi_j \right]}{Z\left[\Lambda\right]}
	\end{equation}
	\begin{equation}
		Z\left[\Lambda\right]
		= \int_{S^2} \exp[\Lambda_{ij} \xi_i \xi_j] d\mathbf{\xi}
	\end{equation}
	Now $p$ is uniquely defined in terms of the Lagrange multipliers $\Lambda_{ij}$.
	Plugging this back into the constraint equation \eqref{eq:Q-def} we get:
	\begin{equation}
	\begin{split}
		Q_{ij} 
		&= \frac{1}{Z[\Lambda]} \left( 
		\int_{S^2} \left( \xi_i \xi_j \exp[\Lambda_{kl} \xi_k \xi_l]
		- \tfrac13 \delta_{ij} \exp[\Lambda_{kl} \xi_k \xi_l] \right)
		d \mathbf{\xi} \right) \\
		&= \frac{1}{Z[\Lambda]} \left(
		\frac{\partial Z[\Lambda]}{\partial \Lambda_{ij}} - \tfrac13 \delta_{ij} Z[\Lambda]
		\right) \\
		&= \frac{\partial \log Z}{\partial \Lambda_{ij}} - \tfrac13 \delta_{ij}
	\end{split}
	\end{equation}
	This set of equations uniquely defines $\Lambda_{ij}$ in terms of $Q_{ij}$, although the equation is not algebraically solvable.
	We may also plug \eqref{eq:p-expr} into \eqref{eq:entropy-def} to get $\Delta S$ as a function of $\Lambda_{ij}$ (and therefore implicitly of $Q_{ij}$):
	\begin{equation}
	\begin{split}
		\Delta S
		&= -n k_B \int_{\Omega} \frac{1}{Z[\Lambda]} \left(
		\int_{S^2} \exp[\Lambda_{ij} \xi_i \xi_j]
		\left(\log(4\pi) + \log(1 / Z[\Lambda]) + \Lambda_{ij} \xi_i \xi_j\right) d \mathbf{\xi}
		\right) d\mathbf{x} \\
		&= -n k_B \int_{\Omega} \left(
		\log(4 \pi) - \log(Z[\Lambda])
		+ \Lambda_{ij} \frac{\partial \log Z[\Lambda]}{\partial \lambda_{ij}}
		\right) \\
		&= -n k_B \int_{\Omega} \left(
		\log(4 \pi) - \log(Z[\Lambda])
		+ \Lambda_{ij} \left( Q_{ij} + \tfrac13	 \delta_{ij} \right)
		\right)
	\end{split}
	\end{equation}
	Further, we may explicitly write out the elastic free energy as:
	\begin{equation}
		f_e (Q_{ij}, \partial_k Q_{ij})
		= L_1 \left(\partial_k Q_{ij}\right) \left(\partial_k Q_{ij}\right)
		+ L_2 \left(\partial_j Q_{ij}\right) \left(\partial_k Q_{ik}\right)
		+ L_3 Q_{kl} \left(\partial_k Q_{ij}\right) \left(\partial_l Q_{ij}\right)
	\end{equation}
	
	\subsection{Finding the equations of motion}
	Now, since $Q_{ij}$ is traceless and symmetric, we need to use a Lagrange multiplier scheme so that there is an extra piece in our free energy:
	\begin{equation}
		f_l = - \lambda Q_{ii} - \lambda_i \epsilon_{ijk} Q_{jk}
	\end{equation}
	To get a time evolution equation for $Q$, we just take the negative variation of the free energy density $f$ with respect to each of them:
	\begin{equation}
	\begin{split}
		\partial_t Q_{ij} 
		= - \frac{\partial f}{\partial Q_{ij}} 
		+ \partial_k \frac{\partial f}{\partial (\partial_k Q_{ij})}
	\end{split}
	\end{equation}
	Let's write out these terms explicitly.
	We start with the Maier-Saupe interaction term:
	\begin{equation}
	\begin{split}
		-\frac{\partial}{\partial Q_{ij}} \left(-\alpha Q_{kl} Q_{lk} \right)
		&= \alpha \delta_{ik} \delta_{jl} Q_{lk}
		+ \alpha \delta_{il} \delta_{jk} Q_{kl} \\
		&= 2 \alpha Q_{ij}
	\end{split}
	\end{equation}
	Now elastic energy:
	\begin{equation}
	\begin{split}
		-\frac{\partial}{\partial Q_{ij}} \left( L_3 Q_{kl} (\partial_k Q_{nm}) (\partial_l Q_{nm}) \right)
		&= - L_3 \delta_{ik} \delta_{jl} (\partial_k Q_{nm}) (\partial_l Q_{nm}) \\
		&= -L_3 (\partial_i Q_{nm}) (\partial_j Q_{nm})
	\end{split}
	\end{equation}
	And the Lagrange multiplier terms:
	\begin{equation}
	\begin{split}
		-\frac{\partial}{\partial Q_{ij}} \left(
		- \lambda Q_{kk} - \lambda_k \epsilon_{klm} Q_{lm}
		\right)
		&= \lambda \delta_{ik} \delta_{jk}
		+ \lambda_k \epsilon_{klm} \delta_{il} \delta_{jm} \\
		&= \lambda \delta_{ij} + \lambda_k \epsilon_{kij}
	\end{split}
	\end{equation}
	Now for the other elastic energy terms:
	\begin{equation}
	\begin{split}
		\partial_k \frac{\partial f}{\partial (\partial_k Q_{ij})} L_1 (\partial_l Q_{nm})(\partial_l Q_{nm})
		&= L_1 \partial_k \left( \delta_{kl} \delta_{in} \delta_{jm} \partial_l Q_{nm}
		+ \partial_l Q_{nm} \delta_{kl} \delta_{ik} \delta_{jm} \right) \\
		&= 2 L_1 \partial_k \partial_k Q_{ij}
	\end{split}
	\end{equation}
	And the $L_2$ term:
	\begin{equation}
	\begin{split}
		\partial_k \frac{\partial f}{\partial (\partial_k Q_{ij})} L_2 (\partial_m Q_{lm}) (\partial_n Q_{ln})
		&= L_2 \partial_k \left( \delta_{km} \delta_{il} \delta_{jm} (\partial_n Q_{ln})
		+ (\partial_m Q_{lm}) \delta_{kn} \delta_{il} \delta_{jn} \right) \\
		&= L_2 \partial_k \left( \delta_{kj} (\partial_n Q_{in}) + \delta_{kj} (\partial_m Q_{im}) \right) \\
		&= 2 L_2 \partial_j (\partial_m Q_{im})
	\end{split}
	\end{equation}
	And finally the $L_3$ term:
	\begin{equation}
	\begin{split}
		\partial_k \frac{\partial f}{\partial (\partial_k Q_{ij})} L_3 Q_{np} (\partial_n Q_{lm}) (\partial_p Q_{lm}) 
		&= L_3 \partial_k Q_{np} \left( \delta_{kn} \delta_{il} \delta_{jm} (\partial_p Q_{lm})
		+ (\partial_n Q_{lm}) \delta_{kp} \delta_{il} \delta_{jm} \right) \\
		&= L_3 \partial_k \left( Q_{kp} (\partial_p Q_{ij}) + Q_{nk} (\partial_n Q_{ij}) \right) \\
		&= 2 L_3 \partial_k \bigl( Q_{kn} (\partial_n Q_{ij}) \bigr)
	\end{split}
	\end{equation}
	Finally, we consider the entropy term:
	\begin{equation}
	\begin{split}
		- \frac{\partial}{\partial Q_{ij}} (-T \Delta s)
		&=
		- \frac{\partial}{\partial Q_{ij}} \left[(-T) (-n k_B) \left(
		\log(4 \pi) - \log( Z[\Lambda] ) + \Lambda_{kl} (Q_{kl} + \tfrac13 \delta_{kl} \right) \right] \\
		&= n k_B T \left(\frac{\partial \log Z}{\partial \Lambda_{kl}} \frac{\partial \Lambda_{kl}}{\partial Q_{ij}}
		- \frac{\partial \Lambda_{kl}}{\partial Q_{ij}} \left( Q_{kl} + \tfrac13 \delta_{kl}\right)
		- \Lambda_{kl} \delta_{ik} \delta_{jl} \right) \\
		&= n k_B T \left( \left( Q_{kl} + \tfrac13 \delta_{kl} \right) \frac{\partial \Lambda_{kl}}{\partial Q_{ij}} 
		- \frac{\partial \Lambda_{kl}}{\partial Q_{ij}} \left( Q_{kl} + \tfrac13 \delta_{kl}\right)
		- \Lambda_{ij} \right) \\
		&= -n k_B T \Lambda_{ij}
	\end{split}
	\end{equation}
	Finally, we need to write down the Lagrange multipliers in terms of $Q$ and its spatial derivatives.
	To do this, note that $Q_{ij}$ is traceless and symmetric so that $\partial_t Q_{ij}$ is also traceless and symmetric.
	Hence, to find $\lambda$ we just take negative $\tfrac13$ the trace of the source term.
	This gives:
	\begin{equation}
	\begin{split}
		\lambda
		&= -\tfrac13 \left(
		- L_3 (\partial_i Q_{nm}) (\partial_i Q_{nm})
		+ 2 L_2 \partial_i ( \partial_m Q_{im} )
		\right) \\
		&= \tfrac13 \left( 
		L_3 (\partial_i Q_{nm}) (\partial_i Q_{nm})
		- 2 L_2 \partial_i ( \partial_m Q_{im} )
		\right)
	\end{split}
	\end{equation}
	where the rest of the terms are traceless.
	Now to find $\lambda_k$, we know that the anti-symmetric piece of any matrix can be given by:
	\begin{equation}
		\tfrac12 \left( A_{ij} - A_{ji} \right)
	\end{equation}
	Further, the Lagrange multiplier term needs to cancel out the anti-symmetric piece:
	\begin{equation}
		\lambda_k \epsilon_{kij} = -\tfrac12 \left( A_{ij} - A_{ji} \right)
	\end{equation}
	To solve for $\lambda_k$ explicitly, we may calculate:
	\begin{equation}
	\begin{split}
		-\tfrac12 \epsilon_{lij} \left( A_{ij} - A_{ji} \right)
		&= \lambda_k \epsilon_{kij} \epsilon_{lij} \\
		&= \lambda_k \left( \delta_{kl} \delta_{ii} - \delta_{ki} \delta_{il} \right) \\
		&= 2 \lambda_l
	\end{split}
	\end{equation}
	Hence:
	\begin{equation}
	\begin{split}
		\lambda_l 
		&= -\tfrac12 L_2 \epsilon_{lij} \left( \partial_j (\partial_m Q_{im}) - \partial_i (\partial_m Q_{jm}) \right) \\
		&= \tfrac12 L_2 \epsilon_{lij} \left( \partial_i (\partial_m Q_{jm})
		- \partial_j (\partial_m Q_{im}) \right)
	\end{split}
	\end{equation}
	since the $L_2$ term is the only one that's anti-symmetric.
	The source term corresponding to this Lagrange multiplier piece is then given by:
	\begin{equation}
	\begin{split}
		\tfrac12 L_2 \bigl( (\partial_k \partial_n Q_{mn}) - (\partial_m \partial_n Q_{kn} ) \bigr) \epsilon_{lkm} \epsilon_{lij}
		&= \tfrac12 L_2 \bigl( (\partial_k \partial_n Q_{mn}) - (\partial_m \partial_n Q_{kn} ) \bigr)
		\left( \delta_{ki} \delta_{mj} - \delta_{kj} \delta_{mi} \right) \\
		&= \tfrac12 L_2 \bigl( (\partial_i \partial_n Q_{jn}) - (\partial_j \partial_n Q_{in} ) \bigr)
		- \tfrac12 L_2 \bigl( (\partial_j \partial_n Q_{in}) - (\partial_i \partial_n Q_{jn} ) \bigr) \\
		&= L_2 \bigl( (\partial_i \partial_n Q_{jn}) - (\partial_j \partial_n Q_{in} ) \bigr)
	\end{split}
	\end{equation}
	Hence, the total equation of motion is:
	\begin{equation}
	\begin{split}
		\partial_t Q_{ij}
		&=
		\begin{multlined}[t]
		2 \alpha Q_{ij}
		- L_3 (\partial_i Q_{nm}) (\partial_j Q_{nm}) 
		+ \tfrac13 \left( L_3 (\partial_k Q_{nm})(\partial_k Q_{nm}) 
		- 2 L_2 (\partial_k \partial_m Q_{km}) \right) \delta_{ij} \\
		+ L_2 \bigl( (\partial_i \partial_n Q_{jn}) - (\partial_j \partial_n Q_{in} ) \bigr)
		- n k_B T \Lambda_{ij} \\
		+ 2 L_1 \partial_k \partial_k Q_{ij} 
		+ 2 L_2 (\partial_j \partial_m Q_{im})
		+ 2 L_3 \partial_k \bigl( Q_{kn} (\partial_n Q_{ij}) \bigr)
		\end{multlined} \\
		&=
		\begin{multlined}[t]
		2 \alpha Q_{ij}
		- L_3 (\partial_i Q_{nm}) (\partial_j Q_{nm})
		- n k_B T \Lambda_{ij} \\
		+ 2 L_1 \partial_k \partial_k Q_{ij} 
		+ L_2 \left( (\partial_j \partial_m Q_{im}) + (\partial_i \partial_m Q_{jm}) \right)
		+ 2 L_3 \partial_k \bigl( Q_{kn} (\partial_n Q_{ij}) \bigr) \\
		+ \tfrac13 \left( L_3 (\partial_k Q_{nm})(\partial_k Q_{nm}) 
		- 2 L_2 (\partial_k \partial_m Q_{km}) \right) \delta_{ij}
		\end{multlined} \\
		&= F_{ij} \left(Q_{ij} ; \partial_k Q_{ij} ; \partial_l \partial_k Q_{ij}\right)
	\end{split}
	\end{equation}
	One can see that $F_{ij}$ is both symmetric and traceless by virtue of $Q_{ij}$ being traceless and symmetric.
	
	\subsection{Reducing degrees of freedom}
	Since $Q_{ij}$ is traceless and symmetric, we only have five independent degrees of freedom.
	We label as follows:
	\begin{equation}
	Q_{ij}
	= \begin{bmatrix}
	Q_1 & Q_2 & Q_3 \\
	Q_2 & Q_4 & Q_5 \\
	Q_3 & Q_5 & -(Q_1 + Q_4)
	\end{bmatrix}
	\end{equation}
	We can define similarly for $F_{ij}$.
	In this case, we just get a five-component vector equation:
	\begin{equation}
		\partial_t Q_i = F_i \left( Q_i ; \partial_j Q_i ; \partial_k \partial_j Q_i \right)
	\end{equation}
	where, we may write $F_i$ as a function of the vector components $Q_i$ (and spatial derivatives thereof) by just explicitly carrying out the sums over the tensor indices.
	This observation reduces the number of equations from 9 down to 5.
	
	\section{Numerically inverting $Q_{ij} (\Lambda)$}
	In all variations of investigating the behavior of nematic-state liquid crystals using a Maier-Saupe free energy, we will need to numerically invert $Q_{ij} (\Lambda)$.
	Hence, we describe the method for doing that first, and implement it as a practice for working with numerical libraries.
	To begin, we have the following explicit expression for $Q_{ij}$ in terms of $\Lambda_{ij}$:
	\begin{equation}
	Q_{ij} (\Lambda)
	= \frac{\int_{S^2} \xi_i \xi_j \exp\left[\Lambda_{kl} \xi_k \xi_l \right] d \xi}
	{\int_{S^2} \exp\left[\Lambda_{kl} \xi_k' \xi_l'\right] d\xi'}
	- \tfrac13 \delta_{ij}
	\end{equation}
	For this we use Newton's method to find the values for $\Lambda$ given fixed values for the components of $Q_{ij}$.
	Note that, during these calculations, we use Lebedev quadrature to do the integrations over the sphere.
	
	To actually implement Newton's method, we do as follows.
	For any fixed $Q_{ij}$ define a vector residual $R_m$ such that:
	\begin{equation}
	R_m (\Lambda)
	= \frac{\int_{S^2} \xi_{i(m)} \xi_{j(m)} \exp\left[\Lambda_{kl} \xi_k \xi_l \right] d \xi}
	{\int_{S^2} \exp\left[\Lambda_{kl} \xi_k' \xi_l'\right] d\xi'}
	- \tfrac13 \delta_{i(m) j(m)}
	- Q_{i(m) j(m)}
	\end{equation}
	where $m$ refers to the index of the distinct degrees of freedom of $Q$, and $(i(m), j(m))$ denotes the location in the matrix.
	We see values of $\Lambda_m$ which will be a zero of this quantity.
	Hence, we must find the Jacobian, which is just:
	\begin{equation}
	R'(\Lambda) = \frac{\partial R_m}{\partial \Lambda_n}
	\end{equation}
	This will be a $5\times 5$ matrix.
	This needs to be done in a non-uniform way because the diagonal elements show up differently than the off-diagonal elements. 
	For the off-diagonal elements we get:
	\begin{equation}
	R'_{mn}
	=
	\begin{multlined}[t]
	\frac{\int_{S^2} 2 \xi_{i(m)} \xi_{i(n)} \xi_{j(m)} \xi_{j(n)}
		\exp\left[\Lambda_{kl} \xi_k \xi_l\right] d\xi}
	{\int_{S^2} \exp\left[\Lambda_{kl} \xi_k \xi_l\right] d\xi} \\
	- \frac{\int_{S^2} \xi_{i(m)} \xi_{j(m)} 
		\exp\left[\Lambda_{kl} \xi_k \xi_l \right]
		\int_{S^2} 2 \xi_{i(n)} \xi_{j(n)}
		\exp\left[\Lambda_{kl} \xi_k \xi_l \right]}
	{\left(\int_{S^2} \exp\left[\Lambda_{kl} \xi_k \xi_l\right] \right)^2}
	\end{multlined}
	\end{equation}
	where the factor of 2 in each term comes from the fact that, for off-diagonal elements, each entry in $\Lambda_n$ appears twice in the $\Lambda_{kl}$ tensor, and the $\xi$ coefficients turn out to be the same (because the expression is symmetric in $k$ and $l$ within the exponent).
	For the diagonal elements, we get:
	\begin{equation}
	R'_{mn}
	= 
	\begin{multlined}[t]
	\frac{\int_{S^2} \xi_{i(m)} \xi_{j(m)} \left(\xi_{i(n)}^2 - \xi_3^2\right)
		\exp\left[\Lambda_{kl} \xi_k \xi_l\right]}
	{\int_{S^2} \exp\left[\Lambda_{kl} \xi_k \xi_l\right] d\xi} \\
	- \frac{\int_{S^2} \xi_{i(m)} \xi_{j(m)}
		\exp\left[\Lambda_{kl} \xi_k \xi_l\right]
		\int_{S^2} \left(\xi_{i(n)}^2 - \xi_3^2\right)
		\exp\left[\Lambda_{kl} \xi_k \xi_l\right]}
	{\left(\int_{S^2} \exp\left[\Lambda_{kl} \xi_k \xi_l \right] \right)^2}
	\end{multlined}
	\end{equation}
	where here the expression in parentheses comes from the fact that the $(3, 3)$ must be such that the tensor is traceless.
	Given these expressions for the Jacobian, we must solve:
	\begin{align}
	R'_{ij} (\Lambda^n) \delta \Lambda^n_j &= -R_m (\Lambda^n) \\
	\Lambda^{n + 1}_{i} &= \Lambda^n_{i} + \delta \Lambda^n_i
	\end{align}
	iteratively until $\left| R_m (\Lambda^n) \right| < \epsilon$ for some small error $\epsilon$.
	
	\section{Isotropic steady state problem}
	As a first problem using finite element analysis, we seek to simplify as much as possible.
	To this end, we only consider the steady-state equation ($\partial_t Q_i = 0$), and we only consider isotropic elasticity ($L_2 = L_3 = 0$).
	Then the governing equation reads:
	\begin{equation}
		0
		= 2 \alpha Q_i
		+ 2 L_1 \partial_k \partial_k Q_i
		- n k_B T \Lambda_i
	\end{equation}
	We may non-dimensionalize by defining a characteristic length-scale:
	\begin{equation}
		\xi = \sqrt{\frac{2 L_1}{n k_B T}}
	\end{equation}
	and then define a dimensionless length and interaction parameter:
	\begin{align}
		x &= \overline{x} \: \xi \\
		\overline{\alpha} &= \frac{2 \alpha}{n k_B T}
	\end{align}
	Dropping the overlines for brevity, our equation of motion reads:
	\begin{equation}
		0 = \alpha Q_i + \partial_k \partial_k Q_i - \Lambda_i
	\end{equation}
	
	\subsection{Numerical scheme}
	Because $\Lambda_i$ is a nonlinear function of $Q_i$, this problem is nonlinear.
	Since the finite element scheme can only handle linear problems, we use Newton's method to solve.
	To that end, we define a residual as exactly the source term:
	\begin{equation}
		F_i (Q)
		= \alpha Q_i + \partial_k \partial_k Q_i - \Lambda_i
	\end{equation}
	We want this to have magnitude zero, and so we iteratively solve the equations:
	\begin{align}
		F'_{ij} (Q^n) \delta Q^n_j &= -F_i(Q^n) \label{eq:isotropic-linear} \\
		Q^{n + 1}_i &= Q^n_i + \delta Q^n_i
	\end{align}
	where $n$ denotes the step in the iteration, and $F'_{ij}$ is the Gateaux derivative of the residual. 
	Writing out this Gateaux derivative gives us:
	\begin{equation}
	\begin{split}
		F'_{ij} \delta Q_{j}
		&= \alpha \frac{\partial Q_i}{\partial Q_j} \delta Q_j
		+ \left(\partial_k \partial_k \frac{\partial Q_i}{\partial Q_j} \delta Q_j \right)
		- \frac{\partial \Lambda_i}{\partial Q_j} \delta Q_j \\
		&= \alpha \delta Q_i
		+ \partial_k \partial_k \delta Q_i
		- \left(\frac{\partial Q_i}{\partial \Lambda_j}\right)^{-1} \delta Q_j \\
		&= \alpha \delta Q + \nabla^2 \delta Q - R'^{-1} (Q) \delta Q
	\end{split}
	\end{equation}
	where in the second to last line we have used the chain rule to rewrite $\partial \Lambda_i / \partial Q_j$ and then in the final line we have written everything in vector notation so that we may use indices to indicate test and basis functions.
	Note that we have used the same Jacobian from our dicussion of inverting $Q(\Lambda)$.
	
	To implement this as a finite element problem, we take \eqref{eq:isotropic-linear} and integrate it against some finite number of test functions enumerated by $i$:
	\begin{equation}
		\left< \phi_i, \alpha \, \delta Q^n \right>
		+ \left< \phi_i, \nabla^2 \delta Q^n \right>
		- \left< \phi_i, R'^{-1}(Q^n) \delta Q^n \right>
		=
		- \left< \phi_i, \alpha Q^n \right>
		- \left< \phi_i, \nabla^2 Q^n \right>
		+ \left< \phi_i, \Lambda(Q^n) \right>
	\end{equation}
	Integrating by parts and taking into account the Dirichlet conditions (so that the test functions go to zero on the boundaries), we get:
	\begin{equation}
		\left< \phi_i, \alpha \, \delta Q^n \right>
		- \left< \nabla \phi_i, \nabla \delta Q^n \right>
		- \left< \phi_i, R'^{-1}(Q^n) \delta Q^n \right>
		=
		- \left< \phi_i, \alpha Q^n \right>
		+ \left< \nabla \phi_i, \nabla Q^n \right>
		+ \left< \phi_i, \Lambda(Q^n) \right>
	\end{equation}
	Now we approximate $\delta Q^n$ as a linear combination of some number of basis functions:
	\begin{equation}
		\delta Q^n = \sum_j \delta Q^n_j \phi_j
	\end{equation}
	Substituting this into our equation yields:
	\begin{equation}
		\sum_j \biggl[
		\alpha \left< \phi_i, \phi_j \right>
		- \left< \nabla \phi_i, \nabla \phi_j \right>
		- \left< \phi_i, R'^{-1}(Q^n) \phi_j \right>
		\biggr] \delta Q^n_j
		=
		- \left< \phi_i, \alpha Q^n \right>
		+ \left< \nabla \phi_i, \nabla Q^n \right>
		+ \left< \phi_i, \Lambda(Q^n) \right>
	\end{equation}
	which is the same as the linear problem:
	\begin{equation}
		A^n_{ij} \delta Q^n_j = b^n_i
	\end{equation}
	with the definitions:
	\begin{equation}
		A^n_{ij} = 
		\alpha \left< \phi_i, \phi_j \right>
		- \left< \nabla \phi_i, \nabla \phi_j \right>
		- \left< \phi_i, R'^{-1}(Q^n) \phi_j \right>
	\end{equation}
	and
	\begin{equation}
		b_i =
		- \left< \phi_i, \alpha Q^n \right>
		+ \left< \nabla \phi_i, \nabla Q^n \right>
		+ \left< \phi_i, \Lambda(Q^n) \right>
	\end{equation}
	
	\section{Numerical scheme}
	\subsection{Time discretization}
	To numerically solve this equation, we use Rothe's method to discretize the time dependence before the spatial dependence.
	To this end, we introduce the following finite difference scheme.
	For $n$ the number of the current time step, call:
	\begin{align}
		k &= t_n - t_{n - 1} \\
		\partial_t Q_i &\to \frac{Q^n_i - Q^{n - 1}_i}{k} \\
		F_i &\to \left[ \theta F_i^n + (1 - \theta) F_i^{n - 1} \right]
	\end{align}
	where $F^n_i$ is just $F_i$ with $Q_i$ evaluated at timestep $n$.
	Here $\theta = 0$ corresponds to an explicit Euler method, while $\theta = 1$ corresponds to an implicit Euler method.
	Also, $\theta = 1/2$ corresponds to a Crank-Nicolson method -- we leave it undefined so that we may play with it later.
	The time-discretized equation is thus:
	\begin{equation}
	\begin{split}
		G_i \left(Q^n_i ; \partial_k Q^n_i ; \partial_l \partial_k Q^n_i\right)
		&= k \left[ \theta F^n_i + (1 - \theta) F^{n - 1}_i \right]
		- Q^n_i + Q^{n - 1}_i \\
		&= 0
	\end{split}
	\end{equation}
	
	\subsection{Space discretization}
	To turn this into a finite element problem, we introduce a scalar residual function:
	\begin{equation}
		R \left(Q^n_i ; \partial_k Q^n_i ; \partial_l \partial_k Q^n_i\right) (\phi_i)
		= \int_\Omega G_i \phi_i
		= 0
	\end{equation}
	where $\phi_i$ is a vector of test functions.
	Now, we would like to only sum over the 5 free components to keep from making redundant calculations.
	However, in the explicit definition of $F_i$ there are two floating indices which are, on some terms, located on differential operators.
	Hence, we consider a tensor test function which is defined in the same way as $Q_{ij}$:
	\begin{equation}
		\phi_{ij}
		= \begin{bmatrix}
		\phi_1 & \phi_2 & \phi_3 \\
		\phi_2 & \phi_4 & \phi_5 \\
		\phi_3 & \phi_5 & -(\phi_1 + \phi_4)
		\end{bmatrix}
	\end{equation}
	We can write the expression for $F^n_i$ out explicitly as follows:
	\begin{equation}
	\begin{split}
		\int_\Omega F^n_i \phi_i
		&=
		\int_\Omega F^n_{ij} \phi_{ij} \\
		&= 
		\begin{multlined}[t]
			2 \alpha \int_\Omega Q^n_{ij} \phi_{ij}
			- L_3 \int_\Omega (\partial_i Q^n_{nm})(\partial_j Q^n_{nm}) \phi_{ij}
			- n k_B T \int_\Omega \Lambda_{ij} \phi_{ij} \\
			+ 2 L_1 \int_{\partial \Omega} n_k (\partial_k Q^n_{ij}) \phi_{ij}
			- 2 L_1 \int_{\Omega} (\partial_k Q^n_{ij})(\partial_k \phi_{ij}) 
			+ L_2 \int_{\partial \Omega} n_m (\partial_j Q^n_{im}) \phi_{ij} \\
			- L_2 \int_\Omega (\partial_j Q^n_{im}) (\partial_m \phi_{ij}) 
			+ L_2 \int_{\partial \Omega} n_m (\partial_i Q^n_{jm}) \phi_{ij}
			- L_2 \int_\Omega (\partial_i Q^n_{jm}) (\partial_m \phi_{ij}) \\
			+ 2 L_3 \int_{\partial \Omega} n_k Q^n_{kn} (\partial_n Q^n_{ij}) \phi_{ij} 
			- 2 L_3 \int_\Omega Q^n_{kn} (\partial_n Q^n_{ij}) (\partial_k \phi_{ij})
			+ \tfrac13 L_3 \int_\Omega (\partial_k Q^n_{nm}) (\partial_k Q^n_{nm}) \phi_{ii} \\
			- \tfrac23 L_2 \int_{\partial \Omega} n_k (\partial_m Q^n_{km}) \phi_{ii}
			+ \tfrac23 L_2 \int_\Omega (\partial_m Q^n_{km}) (\partial_k \phi_{ii})
		\end{multlined}
	\end{split}
	\end{equation}
	where we understand the sum over $(i, j)$ to be only over those components which are distinct in the test function (i.e. (1, 1), (1, 2), (1, 3), (2, 2), (2, 3)), and where $n_i$ is the vector normal to the boundary.
	Note that the other sums (e.g. over $m$) range over all three elements because that is how they appear in the definition of $F_i$.
	
	At this point, we choose Dirichlet boundary conditions so that $\phi_i$ comes from the tangent space and thus takes value zero on the boundary.
	This lets us disregard all of the boundary terms.
	Further, we use the inner product notation $\left< \cdot, \cdot \right>$ to denote the integral over the domain.
	This reduces the expression to:
	\begin{equation}
		\left< F^n_i, \phi_i \right>
		=
		\begin{multlined}[t]
			2 \alpha \left< Q^n_i, \phi_i \right>
			- n k_B T \left< \Lambda_i, \phi_i \right>
			- 2 L_1 \left< \partial_k Q^n_i, \partial_k \phi_i \right> \\
			- L_2 \biggl[
			\left< \partial_j Q^n_{im} + \partial_i Q^n_{jm}, \partial_m \phi_{ij} \right>
			- \tfrac23 \left< \partial_m Q^n_{km}, \partial_k \phi_{ij} \delta_{ij} \right>
			\biggr] \\
			- L_3 \biggl[
			\left< (\partial_i Q^n_{nm}) (\partial_j Q^n_{nm}), \phi_{ij} \right>
			+ 2 \left< Q^n_{kn} (\partial_n Q^n_i), \partial_k \phi_i \right> \\
			- \tfrac13 \left< (\partial_k Q^n_{nm})(\partial_k Q^n_{nm}), \phi_{ij} \delta_{ij} \right>
			\biggr]
		\end{multlined}
	\end{equation}
	With this explicit expression in mind, the residual function is given by:
	\begin{equation}
		R(Q^n) (\phi) =
		k \biggl[
		\theta \left< F^n_i, \phi_i \right> 
		+ (1 - \theta) \left< F^{n - 1}_i, \phi_i \right>
		\biggr]
		- \left< Q^n_i, \phi_i \right>
		+ \left< Q^{n - 1}_i, \phi_i \right>
	\end{equation}
	For each time step, we will need to iteratively solve for $Q^n$ as a zero of this expression using Newton's method.
	Using subscripts to denote the number of the iteration in Newton's method, the method reads:
	\begin{align}
		R'(Q^n_{k - 1}, \delta Q^n_{k - 1}) (\phi)
		&= - R(Q^n_{k - 1}) (\phi) \\
		Q^n_k
		&= Q^n_{k - 1} + \delta Q^n_{k - 1}
	\end{align}
	To be clear about what objects we're dealing with here, for a fixed test function $\phi$, $R(Q^n_{k - 1})$ is a scalar, and $R'(Q^n_{k - 1})$ is a linear operator which acts on $\delta Q^n_{k - 1}$ to produce a scalar.
	In this case, we can create a linear system by solving this equation simultaneously for some number $N$ of basis functions.
	Writing $\delta Q^n_{k - 1}$ as a linear combination of these basis elements, we now have a linear equation in those elements which we may solve for the coefficients.
	Hence, we can find $Q^n$ and thus step forward in time.
	
	\subsection{An example to check}
	Consider a uniaxial, constant $S$ configuration:
	\begin{equation}
		\hat{n} = (\cos\phi, \sin\phi, 0)
	\end{equation}
	\begin{equation}
	\begin{split}
		Q_{ij} 
		&= \frac{S}{2} (3 n_i n_j - \delta_{ij}) \\
		&= \frac{S}{2} 
		\begin{bmatrix}
			3\cos^2\phi - 1 & 3\cos\phi \sin\phi & 0 \\
			3\cos\phi \sin\phi & 3\sin^2\phi - 1 & 0 \\
			0 & 0 & -1
		\end{bmatrix}
	\end{split}
	\end{equation}
	Choose a particularly simple configuration with $\phi = 0$ and $S = 1$ so that we get:
	\begin{equation}
		Q_{ij} = 
		\begin{bmatrix}
			1 &0 &0 \\
			0 &-\tfrac12 &0 \\
			0 &0 &-\tfrac12
		\end{bmatrix}
	\end{equation}
	
	
	\section{A first problem to solve numerically}
	In order to get our feet wet programming with numerical libraries, we begin by solving the problem of numerically inverting $Q_{ij} (\Lambda)$.
	
	
	
\end{document}